Recursive partitioning of linear growth curve models

### 2023-01-20 pkg/glmertree/paper/lgcm-tree/_Partitioning_GCMs_with_GLMM_trees.Rnw
improvements in abstract and intro
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3243 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 22883f9d5661d07da9d7f6258f4d71a06c2b8503

\title{Recursive partitioning of linear growth curve models}
\shorttitle{Recursive partitioning of linear growth curve models}
\authorsnames{Marjolein Fokkema$^1$, Achim Zeileis$^2$}
\authorsaffiliations{{$^1$Leiden University}, {$^2$University of Innbruck}}

\abstract{Growth curve models are popular tools for studying the development of a response variable within subjects over time. Heterogeneity between subjects is likely encountered in such models and researchers are often interested in explaining or predicting this heterogeneity. We show how generalized linear mixed effects model (GLMM) trees can be used to identify heterogeneous subgroups in linear growth curve models. GLMM trees were originally proposed for clustered cross-sectional data; here we extend the method to allow for longitudinal data. The resulting extended GLMM trees are directly applicable to growth curve models as an important special case. We evaluate GLMM tree's performance in partitioning growth curve models through a simulation study and through application to children's development of scholastic abilities. We compare the performance of GLMM trees with three alternative methods: SEM trees, LongCART and longRPart. GLMM trees provide similar accuracy compared to SEM trees, and better accuracy compared to LongCART and longRPart. In addition, GLMM trees allow for modeling both discrete and continuous time series, are less sensitive to (mis-)specification of the random-effects structure and require substantially lower computation times.}

@@

Development over time is of prime interest in psychological researchers. For example, researchers may want to model and predict student's reading or math development in educational studies \citep{MorgyFark09}, or patient's symptom reduction over the course of clinical trails \citep{Fein09}. Mixed-effects or latent-variable models can be used to model development over time, and allow for explaining possible heterogeneity with covariates of a-priori known relevance \citep{NeisyMatt18}. However, the covariates that are predictive of development may often not be known in advance, and identifying them may be of prime interest.

Recursive partitioning methods (RPMs) allow for identifying relevant predictors from a potentially (very) large number of covariates. Several RPMs that allow for partitioning growth-curve models have been developed: GUIDE \citep{Loh02}, longRPart \citep{AbdoyLeBl02}, GEE-based decision trees \citep{Lee05}, longitudinal interaction trees \cite[IT; ][]{SuyMene11}, SEM trees \citep{BranyOert13}, mixed-effects longitudinal trees \citep[MELT; ][]{EoyCho14} and LongCART \cite{KundyHare19} allow for partitioning linear growth curve models (LGCMs). Further, the longRPart2 \citep{StegyJaco18} and IT-LT \citep{WeiyLiu20} methods allow for subgroup detection in non-linear growth curve models\footnote{Both IT methods have a specific aim of detecting subgroups for which the time-by-treatment interaction differs, so are not general methods for partitioning growth curve models.}.

Here we propose the use of generalized linear mixed-effects model trees \citep[GLMM trees; ][]{FokkySmit18} for partitioning LGCMs. GLMM trees were originally proposed for subgroup detection in clustered cross-sectional studies, where subjects are for example nested in treatment centers, classrooms and/or geographical areas. Yet, GLMM trees provide a more general method 2for subgroup detection in almost any type of mixed-effects model. Compared to clustered cross-sectional data, however, longitudinal data may require a different estimation approach: The variance of random effects tends to be higher with longitudinal data, and predictors of interest tend to be measured at higher levels (e.g., time-invariant covariates). In this paper, we propose and test two recent extensions of GLMM trees that account for these characteristics. We focus on the specific use case of partitioning LGCMs, but the extensions are critical for a wide range of settings where covariates are measured at higher levels and/or random-effects variance is substantial.

The main characteristic that sets GLMM trees apart from other methods for partitioning LGCMs is its local-global estimation approach: GLMM trees do not fit a full parametric model in each of the terminal nodes. Instead, the fixed-effects parameters are estimated \textit{locally}, using the observations within a terminal node, and the random-effects parameters are estimated \textit{globally}, using all observations. This local-global estimation approach was first proposed by \cite{HajjyBell11} and \cite{SelaySimo12} for trees with constant fits in the terminal nodes. With GLMM trees, the approach was generalized to allow for GLMs with any number of parameters in the terminal nodes, thus allowing for non-Gaussian responses and targeted detection of a wide range of possible interaction effects \cite{FokkySmit18}.

All other methods for partitioning LGCMs take a fully local estimation approach: Within every terminal node, a full parametric model is estimated based the observations in that node only. This fully local estimation approach brings higher computational load and increased model complexity. GLMM trees require estimation of a (much) lower number of random-effects parameters, which likely reduces overfitting and improves stability and generalizability of the fitted model. Furthermore, the fully local estimation approach requires possible partitioning variables to be measured at the highest level of the multilevel structure, while the local-global estimation approach allows for partitioning variables to be measured at any level.

The computational advantage of GLMM trees may be highest compared to longRPart, longRPart2, IT-LT and LRT-based SEM trees. These methods employ an exhaustive split detection procedure: For every possible split in the current node, the full parametric model is re-estimated in the two nodes that would result from the split. To choose the optimal split, a likelihood ratio test is computed for every possible cut point. Not only does this bring a heavy computational load, it also introduces a selection bias towards covariates with a larger number of possible cut points \citep{ShihyTsai04, Shih04}. LongCART, MELT, GEE-based decision trees and score-based SEM trees also fit full parametric models in each of the nodes, but do not refit models for cutpoint selection; they employ the predictions or residuals from the fitted model in the current node for selecting the best split. This not only reduces computational load, it also allows for separating variable and cutpoint selection, thus preventing selection bias. The GLMM tree algorithm also employs a two-step approach to split selection.

@@

In what follows, we first describe estimation of GLMM trees, and possible adjustments to better account for longitudinal data structures. Next, we assess and compare performance of GLMM trees for partitioning LGCMs: In Simulation~Study~I, we assess performance of the proposed adaptations to GLMM trees. In Simulation~Study~II, we compare performance of GLMM trees with that of two other methods: SEM trees and LongCART. In the Application, we compare the performance of GLMM trees and longRPart in existing datasets on children's development of reading, math and science abilities. In the Discussion, we summarize our findings and implications.

@@

In the GLMM tree model \citep{FokkySmit18}, expectation $\mu_i$ of outcome vector $y_i$ is modeled through a linear predictor and suitable link function:
%
\begin{eqnarray}
\label{eq:expected_value}
E[y_i | X_i] & = & \mu_i, \\
\label{eq:GLMMtree}
g(\mu_{i}) & = & X_{i} \beta_{j} + Z_{i} b_{i}
\end{eqnarray}
%
Thus, in the current paper, $g$ is the identity function and we assume a continuous response with normally distributed errors and constant variance $\sigma_\epsilon$. Generalizations to other response variable types within the GLM are straightforward. Further, $X_i$ is the $n_i \times (p+1)$ fixed-effects design matrix for subject $i$ ($i = {1, \dots, N}$). We furter assume time is the predictor variable of interest, thus $p=1$ and $X_i$ comprises a column of 1s for the intercept and a column of observed timepoints for subject $i$. Note that the number and spacing of observed timepoints may differ between subjects. $\beta_j$ is a column vector of node-specific fixed-effects parameters, its value depending on terminal node $j$ of which observation $i$ is part. As in a traditional GLMM, $Z_i$ is the random-effects design matrix for subject $i$, comprising a subset of columns of $X_i$, and $b_i$ is a vector of random effects for subject $i$. Finally, $b$ is assumed to follow a (possibly multivariate) normal distribution with mean zero and (co)variance $\Sigma_{b}$.

>>>>

\title{Recursive partitioning of linear growth curve models}
\shorttitle{Recursive partitioning of linear growth curve models}
\authorsnames{Marjolein Fokkema$^1$, Achim Zeileis$^2$}
\authorsaffiliations{{$^1$Leiden University}, {$^2$Universit\"at Innsbruck}}

\abstract{Growth curve models are popular tools for studying the development of a response variable within subjects over time. Heterogeneity between subjects is often encountered in such models and researchers are typically interested in explaining or predicting this heterogeneity. We show how generalized linear mixed effects model (GLMM) trees can be used to identify subgroups associated with different sets of parameters in linear growth curve models. Originally developed for clustered cross-sectional data, GLMM trees are extended here to longitudinal data. The resulting extended GLMM trees are directly applicable to growth curve models as an important special case. We evaluate GLMM tree's performance in partitioning growth curve models through a simulation study and through application to children's development of scholastic abilities. We compare the performance of GLMM trees with three alternative methods: structural equation model (SEM) trees, LongCART, and longRPart. GLMM trees provide similar accuracy compared to SEM trees, and better accuracy compared to LongCART and longRPart. In addition, GLMM trees allow for modeling both discrete and continuous time series, are less sensitive to (mis-)specification of the random-effects structure and require substantially lower computation times.}

@@

Development over time is of prime interest in psychological research. For example, researchers may want to model and predict student's reading or math development in educational studies \citep{MorgyFark09}, or patient's symptom reduction over the course of clinical trails \citep{Fein09}. Mixed-effects or latent-variable models can be used to model development over time, and allow for explaining possible heterogeneity with covariates of a-priori known relevance \citep{NeisyMatt18}. However, the covariates that are predictive of development may often not be known in advance, and methods for identifying them are needed.

Recursive partitioning methods (RPMs), also known as ``trees``, allow for identifying relevant predictors from a potentially (very) large number of covariates. Several RPMs that allow for partitioning growth-curve models have been developed: GUIDE \citep{Loh02}, longRPart \citep{AbdoyLeBl02}, GEE-based decision trees \citep{Lee05}, longitudinal interaction trees \cite[IT; ][]{SuyMene11}, SEM trees \citep{BranyOert13}, mixed-effects longitudinal trees \citep[MELT; ][]{EoyCho14} and LongCART \cite{KundyHare19} allow for partitioning linear growth curve models (LGCMs). Further, the longRPart2 \citep{StegyJaco18} and IT-LT \citep{WeiyLiu20} methods allow for subgroup detection in non-linear growth curve models\footnote{Both IT methods have a specific aim of detecting subgroups for which the time-by-treatment interaction differs, so are not general methods for partitioning growth curve models.}.

Here we extend generalized linear mixed-effects model trees \citep[GLMM trees; ][]{FokkySmit18} so that they can be applied to partitioning LGCMs. GLMM trees were originally proposed for subgroup detection in clustered cross-sectional studies, where subjects are for example nested in treatment centers, classrooms and/or geographical areas. Moreover, the general idea of GLMM trees is also appealing for subgroup detection in almost any type of mixed-effects model. Compared to clustered cross-sectional data, however, longitudinal data may require a different estimation approach: The variance of random effects tends to be higher with longitudinal data, and predictors of interest tend to be measured at higher levels (e.g., time-invariant covariates). In this paper, we propose and test two extensions of GLMM trees that account for these characteristics. We focus on the specific use case of partitioning LGCMs, but the extensions are also critical for a wide range of further settings where covariates are measured at higher levels and/or where random-effects variance is substantial.

The main characteristic that sets GLMM trees apart from other methods for partitioning LGCMs is its local-global estimation approach: GLMM trees do not fit a full parametric model in each of the terminal nodes. Instead, the fixed-effects parameters are estimated \textit{locally}, using the observations within a terminal node, and the random-effects parameters are estimated \textit{globally}, using all observations. This local-global estimation approach was first proposed by \cite{HajjyBell11} and \cite{SelaySimo12} for trees with constant fits (i.e., intercepts only) in the terminal nodes. With GLMM trees, the approach was generalized to allow for GLMs with any number of parameters in the terminal nodes, thus allowing for non-Gaussian responses and targeted detection of a wide range of possible interaction effects \citep{FokkySmit18}.

All other methods for partitioning LGCMs take a fully local estimation approach: Within every terminal node, a full parametric model is estimated based on the observations in that node only. This fully local estimation approach provides more flexibility but also comes at a higher computational burden and increased model complexity. GLMM trees require estimation of a (much) lower number of random-effects parameters, which likely reduces overfitting and improves stability and generalizability of the fitted model. Furthermore, the fully local estimation approach requires possible partitioning variables to be measured at the highest level of the multilevel structure, while the local-global estimation approach allows for partitioning variables to be measured at any level.

The computational advantage of GLMM trees may be highest compared to longRPart, longRPart2, IT-LT and SEM trees based on likelihood ratio tests (LRTs). All of these methods employ an exhaustive split detection procedure: For every possible split in every available partitioning variable in the current node, the full parametric model is re-estimated in the two nodes that would result from the split. To choose the optimal split, some split criterion (such as a $p$-value from the corresponding LRT) is computed for every possible cutpoint. Not only does this bring a heavy computational load, it also introduces a selection bias towards covariates with a larger number of possible cutpoints \citep{ShihyTsai04, Shih04}. LongCART, MELT, GEE-based decision trees and score-based SEM trees also fit full parametric models in each of the nodes, but do not refit models for cutpoint selection; they employ the predictions or residuals from the fitted model in the current node for selecting the best split. This not only reduces computational load, it also allows for separating variable and cutpoint selection, thus preventing selection bias. The GLMM tree algorithm also employs a two-step approach to split selection.

@@

In what follows, we first describe estimation of GLMM trees in general along with the novel adjustments geared towards longitudinal data structures. Subsequently, we assess and compare performance of GLMM trees for partitioning LGCMs: In Simulation~Study~I, we assess performance of the proposed adaptations to GLMM trees. In Simulation~Study~II, we compare performance of GLMM trees with that of two other methods: SEM trees and LongCART. In the Application, we compare the performance of GLMM trees and longRPart in existing datasets on children's development of reading, math and science abilities. In the Discussion, we summarize our findings and implications.

@@

In the GLMM tree model \citep{FokkySmit18}, expectation $\mu_i$ of outcome vector $y_i$ is modeled through a linear predictor and suitable link function:
%
\begin{eqnarray}
\label{eq:expected_value}
E[y_i | X_i] & = & \mu_i, \\
\label{eq:GLMMtree}
g(\mu_{i}) & = & X_{i} \beta_{j} + Z_{i} b_{i}
\end{eqnarray}
%
Throughout this paper we focus on the case with a continuous response with normally distributed errors and constant variance $\sigma_\epsilon$. Therefore, the identity function can be used for the link $g$. Generalizations to other response variable types within the GLM are straightforward. Further, $X_i$ is the $n_i \times (p+1)$ fixed-effects design matrix for subject $i$ ($i = {1, \dots, N}$). We further assume time is the predictor variable of interest, thus $p=1$ and $X_i$ comprises a column of 1s for the intercept and a column of observed timepoints for subject $i$. Note that the number and spacing of observed timepoints may differ between subjects. $\beta_j$ is a column vector of node-specific fixed-effects parameters, its value depending on terminal node $j$ of which observation $i$ is part. As in a traditional GLMM, $Z_i$ is the random-effects design matrix for subject $i$, comprising a subset of columns of $X_i$, and $b_i$ is a vector of random effects for subject $i$. Finally, $b$ is assumed to follow a (possibly multivariate) normal distribution with mean zero and (co)variance $\Sigma_{b}$.
